## Working with vectors

You can create vectors (an ordered list of elements) by combining elements with the `c` function.

```{r}
first.vector <- c(38, 32, 45, 23, 6, 56,
                  234, 1239, 123, 123, 342, 1234)
second.vector <- c(1, 34, 10)
first.vector
second.vector
```

```{r}
third.vector <- c(first.vector, second.vector)
third.vector
fruits <- c("apples", "oranges", "bananas")
```

R comes with pre-built variables that you can use, such as the vector of all alphabetic characters.

```{r}
letters
LETTERS
```

There are several helpful functions and an operator that can speed up your ability to generate vectors of numeric values. Explore what the `:` operator does below and generate a few numeric vectors of your own.

```{r}
1:4
```

```{r,eval=FALSE}
?seq
?rep
```

```{r}
z <- seq(-pi, pi, length = 20)
z
```

```{r}
x <- c(1, 3, 2)
x
```

```{r}
y <- c(1, 4, 3)
length(x)
length(y)
```

What does the `length` function do? Test it on a few more vectors that you have created.

```{r}
x + y
```

You can perform mathematical operations on an entire numeric vector all at once.

```{r}
first.vector * 2
```

```{r,eval=FALSE}
?sin
```

```{r}
sin(first.vector)
```

------

### Indexing and subsetting vectors

Indexing is a way to access into a vector (or matrix/data frame) and pull out certain elements. There are multiple ways to index into a vector, one of the easiest ways is to pull out an element based on its order/position in the grouping (its index). We use `[]` immediately after the name of the grouping (in this case a vector) to access into it.

```{r}
second.vector[2] # pull out second element
fruits[3] # pull out third fruit
```

```{r}
fruits[2:3] # pull out 2nd and 3rd fruit
fruits[c(1, 3)] # pull out 1st and 3rd fruit
```

You can remove elements of a vector by using the same syntax of indexing, but instead put a negative sign in front of the index number.

```{r}
fruits[-1]
```

------

### Using logic and logicals

You can also access elements in a vector that meet certain criteria (using conditional logic).

```{r}
second.vector < 20
second.vector[second.vector < 20]
```

You can use the standard comparison operators like `>` or `<` to check greater than or less than. You can also use `==` to check for equality or `!=` to check that values are not equal.

```{r}
save.num <- 7
save.num + 10
save.num != 8
save.num == 8
```

Conditional logic is where a statement evaluates to True or False. You can pair these conditional statements with an `if`/`else` statement that breaks up the code and only executes parts of the code where conditions are met.

```{r}
test.num <- 30

if (test.num < 10) {
  print("small")
} else {
  print("big")
}
```
