## Working with vectors

You can create vectors (an ordered list of elements) by combining elements with the `c` function.

Let us pretend that we measured and recorded the resting heart rates of several patients.

```{r}
heart.rates <- c(78, 68, 95, 82, 69, 63, 86, 74, 64, 62)
more.heart.rates <- c(86, 79, 64, 74, 80)
heart.rates
more.heart.rates
```

```{r}
all.heart.rates <- c(heart.rates, more.heart.rates)
all.heart.rates
patient.names <- c("oakley", "rashmi", "kiran",
                   "eun", "sasha", "mattie")
```

R comes with pre-built variables that you can use, such as the vector of all alphabetic characters.

```{r}
letters
LETTERS
```

There are several helpful functions and an operator that can speed up your ability to generate vectors of numeric values. Explore what the `:` operator does below and generate a few numeric vectors of your own.

```{r}
1:4
```

```{r,eval=FALSE}
?seq
?rep
```

```{r}
z <- seq(-pi, pi, length = 20)
z
```

```{r}
x <- c(1, 3, 2)
x
```

```{r}
y <- c(1, 4, 3)
length(x)
length(y)
```

What does the `length` function do? Test it on a few more vectors that you have created.

```{r}
x + y
```

You can perform mathematical operations on an entire numeric vector all at once.

```{r}
all.heart.rates + 1
```

```{r,eval=FALSE}
?log
```

```{r}
log(all.heart.rates)
```

```{r,eval=FALSE}
?signif
```

```{r}
signif(all.heart.rates, digits = 1)
```

------

### Indexing and subsetting vectors

Indexing is a way to access into a vector (or matrix/data frame) and pull out certain elements. There are multiple ways to index into a vector, one of the easiest ways is to pull out an element based on its order/position in the grouping (its index). We use `[]` immediately after the name of the grouping (in this case a vector) to access into it.

```{r}
all.heart.rates[2] # pull out second heart rate measurement
patient.names[3] # pull out third patient name
```

```{r}
patient.names[2:3] # pull out 2nd and 3rd patient name
patient.names[c(1, 3)] # pull out 1st and 3rd patient name
```

You can remove elements of a vector by using the same syntax of indexing, but instead put a negative sign in front of the index number.

```{r}
patient.names[-1]
```

------

### Using logic and logicals

You can also access elements in a vector that meet certain criteria (using conditional logic).

```{r}
all.heart.rates < 80
all.heart.rates[all.heart.rates < 80]
```

You can use the standard comparison operators like `>` or `<` to check greater than or less than. You can also use `==` to check for equality or `!=` to check that values are not equal.

```{r}
save.num <- 7
save.num != 8
save.num == 8
```

Conditional logic is where a statement evaluates to True or False. You can pair these conditional statements with an `if`/`else` statement that breaks up the code and only executes parts of the code where conditions are met.

```{r}
test.num <- 30

if (test.num < 10) {
  print("small")
} else {
  print("big")
}
```
