## Reading and processing data


find a file-reading function, know of a few
use read.csv to read a file and save it as a variable, and tune some arguments

```{r, eval=FALSE}
library(tidyverse)
```

```{r}
irisz <- read.csv("data/iris.csv")
head(irisz)
```

```{r}
head(irisz$Species)
```

Settings for read.csv

```{r}
irisz <- read.csv("data/iris.csv", stringsAsFactors = TRUE)
head(irisz$Species)
head(irisz$Location)
```

```{r}
irisz <- read.csv("data/iris.csv", col.names = c("sep_len", "sep_wid", "pet_len", "pet_wid", "species", "loc"))
head(irisz)
```

important to preview data and look for issues, and have some strategies/functions to do this

Examining Data

```{r}
str(irisz)
summary(irisz)
```

```{r}
unique(irisz$sep_len)
```

what NAs are, some strategies to deal with these

Handling Missing Data

```{r}
irisz <- read.csv("data/iris.csv", na.strings = c("", "n/a"), stringsAsFactors = TRUE)
head(irisz)
```

```{r}
str(irisz)
summary(irisz)
```

Looking for anamolous data

```{r}
hist(irisz$Sepal.Length)
hist(irisz$Sepal.Width)
```

```{r}
hist(irisz$Petal.Length)
hist(irisz$Petal.Width)
```

how to filter entries in a data.frame


how to cast/transform value Types
how to scale/normalize values

```{r}
head(irisz$Sepal.Length)
```

```{r, eval = FALSE}
?scale
```

```{r}
x <- 1:10
scale(x)
scale(x, center = FALSE)
scale(x, scale = FALSE)
```

```{r}
head(scale(irisz$Sepal.Length))
```

```{r}
library(scales)
rescale(x)        
```

