## Making plots with ggplot2

**Learning Goals:**

* List the roles of data, aesthetics, and geoms in ggplot functions.
* Choose the correct aesthetics and alter the geom parameters for a scatter plot, histogram, or box plot.
* Layer multiple geometries in a single plot.
* Customize plot scales, titles, subtitles, themes, fonts, layout, and orientation.
* Apply a facet to a plot.
* Save a ggplot to a file.

### Getting started with a ggplot

```{r,ggplot2,message=FALSE}
library(ggplot2)
library(tidyverse)
```

We will primarily be working in ggplot2 as it has the greatest degree of customization for visualization and offers many additional features over the basic plotting in R.

Most ggplot calls to create a figure take the following form (you can read more using `help(ggplot)`):

`ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()`

We will practice using our mammalian sleep dataset. You can look up more info about this dataset using the `help` function and the dataset name, `msleep`.

```{r}
head(msleep)
```

You will first use the `ggplot()` function and bind the plot to a specific data frame using the data argument.

`ggplot(data = msleep)`

You will next need to define a mapping (using the aesthetic or `aes` function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

`ggplot(data = msleep, aes(x = brainwt, y = sleep_rem))`

You can then add ‘geoms’ – graphical representations of the data in the plot (points, lines, bars). ggplot2 offers many different geoms. We will use some common ones today including:

* `geom_point()` for scatter plots, dot plots, etc.
* `geom_boxplot()` for, well, boxplots!
* `geom_line()` for trend lines, time series, etc.

To add a geom to the plot use the `+` operator. Because we have two continuous variables, let’s use `geom_point()` first:

`ggplot(data = msleep, aes(x = brainwt, y = sleep_rem)) + geom_point()`

To save your work-in-progress, you can assign the plot to a variable.

```{r}
my_plot <- ggplot(data = msleep, aes(x = brainwt, y = sleep_rem))
```

We can now draw the plot as a scatterplot with points to represent each state.

```{r}
my_plot + geom_point()
```

You might notice that all of the points are squished against the y-axis since many of the mammals in this dataset have low brain weights. You can see this if you plot a very simple histogram of `brainwt` to see its distribution, or if you get summary statistics.

```{r}
hist(msleep$brainwt)
summary(msleep$brainwt)
```

As you can see, the minimum and median values are very low, but there are a few mammals with high `brainwt` as you can see by the much larger maximum value in this vector.

To make more useful plots, we can transform this value using log-scaling. While we will have to note that the new values do not exactly match the real-world measurements anymore, patterns we see (i.e. something that correlates with higher brain weights) will still hold true.

```{r}
msleep2 <- msleep %>% mutate(brainwt_log = log(brainwt))
ggplot(msleep2, aes(x = brainwt_log, y = sleep_rem)) + geom_point()
```

Here we use the `mutate` function to make a new variable called `brainwt_log` in our dataset (technically a new dataset copy that we have saved as `msleep2`). Plotting this variable as our x variable (i.e. independent variable), makes it easier to look for patterns.

------

### Plot Aesthetics

???

We can modify the appearance of the plot by changing aspects of the points we plot such as transparency ("alpha") and color.

```{r}
ggplot(data = msleep2, aes(x = brainwt_log, y = sleep_rem)) +
    geom_point(alpha = 0.5, color = "blue")
```

It is also possible to scale the color of the points by some variable present in the data.

```{r}
ggplot(msleep2, aes(x = brainwt_log, y = sleep_rem, color = vore)) + 
  geom_point()
```

When generating visualizations, it is important to annotate the figure with meaningful labels on the axes to make them accessible for the viewer.

```{r}
ggplot(msleep2, aes(x = brainwt_log, y = sleep_rem, color = vore)) + 
  geom_point() + 
  labs(x = "Brain Weight (log)", 
       y = "Duration of REM Sleep")
```
