---
output:
  pdf_document: default
  html_document: default
---

## Tidyverse 1/2: functions & pipes

Goal: Understand the tidyverse framework of pipes and functions, as well as how to apply it to tidy data frame cleaning and manipulation.

### Welcome to the tidyverse! 

[Tidyverse](https://www.tidyverse.org) is a suite of R packages that are very useful for data science when we need to manipulate data frames and visualize data. The most important packages in tidyverse for our workshop will be [dplyr](https://dplyr.tidyverse.org) and [ggplot2](https://ggplot2.tidyverse.org). It builds on the base R functions and data types we've studied so far. It just provides a different design framework for working with data in R. The main two features of tidyverse are 

1. A new suite of very useful, intuitive functions
2. The pipe `%>%` which will allow us to go send data from one function to the next without using intermediate steps or nesting functions together

#### Preview: new functions in tidyverse

Let's say we have a dataframe, `df`, which has the name and age (in years) of everyone in a class. We want to create a new variable of age in months. In base R, we would do it this way:

`df$age_months <- df$age_years * 12`

In tidyverse, we can use the function `mutate()` to create a new variable. As with most tidyverse functions, `mutate()` takes the dataset as the first argument and the operation as the second. Here the operation is to create the new variable.

`mutate(df, age_months = age_years * 12)`

As you will see with more complex examples, **these functions can make dataset operations more readable, especially when strung together with a pipe**. 

#### Preview: the pipe in tidyverse

Recall the `head()` function used in the previous section. If you have a data frame, `df`, instead of peaking at the first few lines with `head(df)`, you can pipe it into the `head()` function like this: `df %>% head()`. This might seem like more work than just calling the function around the data frame object, but when you need to run a bunch of functions on a data frame, it becomes much TIDIER! 

In the previous example, we could call the function using a pipe like so:

`df %>% mutate(age_months = age_years * 12)`


Feeling motivated to accelerate data cleaning and analysis using the tidyverse?


### Load tidyverse

Load the tidyverse library

```{r,loadtidy, message=FALSE}
library(tidyverse)
```

If tidyverse is already installed, loading the library should not return an error message. It is ok if it says there are package conflicts - this happens when different packages that are loaded have functions with the same name (but probably do different things). Just take note of the conflicts. If you use those other packages/functions then you'll have to specify which package the function you're using comes from. For example if `mutate()` was also in another package, to use the tidyverse version I would write `tidyr::mutate()` instead of just `mutate()`

If loading the library returns an error message saying it's not installed, then go to the [SSRP Pre-installation guide](https://tinyurl.com/ssrpsoftwareguide) and follow the instructions for installing tidyverse

### How to pipe %>%

The tidyverse pipe can be used with most base R functions and with all tidyverse-specific functions, which we'll learn more about soon. Although we'll focus on using tidyverse pipe and functions on data frames, the pipe can also be used on vectors and even scalars.


##### Pipes with dataframes

Let's read in the iris dataset used in the previous section. Peak at it with `head()` to check that the columns look intact.

```{r t1_pipes, cache = T}

# Load the dataset and peak at it with the head() function
irisz <- read.csv("data/iris.csv", 
                  na.strings = c("", "n/a"), 
                  stringsAsFactors = TRUE)

head(irisz)

```

Now let's try this using the pipe `%>%`.

```{r t1_pipes2, cache = T}

irisz %>% head()

```

Practice this with other functions we've used so far: `summary()`, `unique()`, `str()`, `dim()`, `table()`, and even `hist()`!. Do it with and without the pipe.

Pretty simple right? 

Notice how the `%>%` pipe is sending the output from before it to the function after it. This `%>%` is similar to the `|` we use on the command line to send data from one operation to the next: `sort data.txt | uniq -c` (sort a data file and count the number of unique lines)

**Pipes can string together functions**

In the previous section, we learned how to read in a file with `read.csv()`, get unique values with `unique()`, and get the dimensions of a data frame with `dim()`. Let's try these operations using the pipe instead of performing them as separate operations.

```{r t2_pipe3, cache = T}

irisz %>%
  unique() %>%
  dim()

```

Notice how each operation is on one line so it's easy to read the sequence of operations. 

**Pipes with scalars**

To take the square root of a number `x` and round it to `y` digits, we would use the `sqrt()` and `round()` functions

Here's what this looks like with and without the pipe:

```{r t2_pipe4, cache = T}

x = 112
y = 2

# Two steps without the pipe
z <- sqrt(x)
round(z, digits = y)

# One step without the pipe
round(sqrt(x), digits = y)

# One step with the pipe
sqrt(x) %>% round(digits = y)

```


**Pipes with vectors**

We can apply the same piping method to vectors. Think about the vector that is the column `Sepal.Length` in the dataframe `irisz` which we have been using.

We'll use pipes to string together the following operations starting with the **vector** `irisz$Sepal.Length`:  sort it, make the values numeric, and make a histogram. (These three functions were also introduced in the previous section.)

```{r t2_pipe5, cache = T}

# One step with the pipe
irisz$Sepal.Length %>%
  as.numeric() %>%
  sort() %>%
  hist()

```
```{r t2_hist2, cache = T, eval = F}

# One step without the pipe requires lots of parentheses
hist(sort(as.numeric(irisz$Sepal.Length)))

```


Hopefully by now you have a sense for how the pipe in tidyverse works and how it can work with functions we already know and can be applied to data frames, vectors, and scalars. Another big contribution tidyverse makes to the data science in R is the addition of new functions!

### Essential tidyverse functions

Let's start with the essential functions of `tidyverse`. All of these are from the sub-package [dplyr](https://dplyr.tidyverse.org)

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `arrange()`: sort results
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `count()`: count discrete values

**Reminder**: if you ever want to know more about these functions, just use the help function - `?` followed by the function of interest. For example: `?select`

#### Selecting, deleting, reordering columns

**Selecting columns**

In base R, if we wanted to select certain columns (e.g. column names c1, c2, c6, c7) from a dataframe, `df`, we would do something like this: `df[c("c1", "c2", "c6", "c7")]`. With tidyverse, we can use the `select(c1, c2, c6, c7)` function (from `dplyr`).

**Deleting columns**

To delete a column, just add a `-` before the column name inside of `select()`.

**Reordering columns**

Select can also be used to reorder columns. Just specify the order of columns in `select()`.

**Selecting and deselecting in the `irisz` dataset**

```{r select_dplyr, eval = F, cache = TRUE}

# Let's select just the location 
select(irisz, Location)

# Remove the Location column
select(irisz, -Location)

```


**Selecting multiple columns**

Let's select both the Location and Species columns and see what unique combinations are present in the dataset

```{r select_twovar, cache = T}

unique(select(irisz, c(Location, Species)))

# Note that the order returned is the order specified in select()
# Originally Location came after Species, but now they are reordered.

```

We can also do this in with the pipe. 

```{r withpipe, eval = T}

irisz %>%
  select(Location, Species) %>%
  unique()

```

Note that `select()` subsets the dataframe to the specified columns. The output is a **data frame** (not a vector), even if only one column is selected

**What if we want to extract one column as a vector? **

In base R, we would use the `$` operation to select a single column as a vector. With tidyverse, we can use the `pull()` function (from `dplyr`). Notice that pull() will only accept a single column name and will return a vector

```{r pull, eval = F}

# Base R
irisz$Location

# With the pipe and tidyverse function pull()
irisz %>%
  pull(Location) 

```

#### Filtering dataframes

The basics of filtering data frames in tidyverse are the `filter()` function and boolean operators (`>`, `<`, `==`) learned in section 2.4.3. 
Let's try two examples:

```{r filter_df,cache = TRUE}

# Filter for Petal.Length greater than x

x = 7

irisz %>%
  filter(as.numeric(Sepal.Length) > x)

# Filter to get observations of setosa species from Korea

irisz %>%
  filter(Species == "setosa",
         Location == "Korea")

```


#### Creating and renaming columns

Creating columns in tidyverse is simple with the `mutate()` function.

Let's say we wanted to create a new column for the `irisz` dataset which has the row number.

In base R, we would do something like `irisz$row_number <- seq(1,nrow(irisz))`. Here, `seq()` creates a sequence of numbers from 1 to the number of rows in `irisz`.

In tidyverse we can do it using the `mutate()` and `row_number()` functions. So it doesn't output the whole data frame, we'll pipe it directly into `head()`

```{r createcol, cache = T}

irisz %>% 
  mutate(row_number = row_number()) %>%
  head()

```
See the row number column?

Mutate can also be used to update a column without creating a new one.

```{r}

# Change sepal length to be 10x what it currently is

irisz %>% 
  mutate(Sepal.Length = 10*Sepal.Length ) %>%
  head()


```

What if all the setosa measurements were actually supposed to be the virginica species?
Let's use the `ifelse()` function which accepts a conditional test, value if true, value if false.

```{r t2_ifelse}

irisz %>% 
  # If Species is setosa, change it to virginica, otherwise use the value in Species 
  mutate(Species = ifelse(Species == "setosa", "virginica", Species )) %>%
  head()


```

**Renaming columns**

Sometimes, we just need to rename an existing column. Use `rename()` to specify the new and old column names, `rename(new_name = old_name)`. Let's rename the location column to be "Country". Let's just pipe it directly into `colnames()` which will give us the column names of the data frame.

```{r t2_rename, eval = T}

irisz %>% 
  rename(Country = Location) %>%
  colnames()

```


Want to change row names into a column or a column into a row name? Check out `rownames_to_column()` and `column_to_rownames()`.

#### Summarizing data

Summarizing data is very important in data analysis. It allows you to get a big picture overview of a new dataset: type, range of values, etc. These are also types of summaries to think about including when in the exploratory stages or a project: before running analyses on a dataset, create a summary of the dataset. Present this during your update meetings for your summer project and keep these summaries on hand in case you or others have questions about the dataset throughout the course of your project.

For summarizing data in tidyverse, let's explore the functions 

- `count()`
- `group_by()`
- `summarize()`

The function `count()` does what it sounds like - it counts the number of unique observations for the specified column(s)

**Count on one variable in the data frame**
```{r count1, cache = TRUE}

irisz %>%
  count(Location)

```

**Count on more than one variable in the data frame**

```{r count2, cache = TRUE}

# Count the number of observations for each combination
# of location and species

irisz %>%
  count(Location, Species)

```

Can you think of another (perhaps longer) way to count using other functions, such as `select()`, `unique()`, `n()`, `pull()`, or `table()`?

**Summarizing WITHIN groups**

What if we want to do more than just count the number of occurrences for a given group? `group_by` takes a variable and groups the data by that variable. Anything that happens after that is only calculated within the group variable. To turn off grouping, use `ungroup()`. It's always good practice to `ungroup()` after you're done grouping. Although it doesn't matter if you won't be doing any more operations on the dataset.

```{r t2_group_by, eval = T, cache = TRUE}

# Here's an example that produces a similar result 
# as in the count() example above

irisz %>%
  group_by(Location) %>%
  summarize(n = n()) %>%
  ungroup()

```
Try taking away the `ungroup()` and see if it changes the result.


Let's build on this. We also want to calculate the `mean()` and `sd()` for a new variable called size for each location. Size is going to be `length * width` of the sepal. Our process is:

1. Create a new column for size using `mutate()`
2. Group the data by Location using `group_by()`
3. Summarize the mean and standard deviation of the group using `summarize()`

```{r t2_group_by2, cache = T}

# Since mean() and sd() require numeric inputs,
# We'll also tell mean() and sd() to ignore NAs in the calculation

irisz %>%
  mutate(size = as.numeric(Sepal.Length) * as.numeric(Sepal.Width)) %>%
  group_by(Location) %>%
  summarize(n = n(),
            mean_size = mean(Petal.Length, na.rm = T),
            mean_sd = sd(Petal.Length, na.rm = T)) %>%
    ungroup()

```

For a further challenge, let's clean up this table. Arrange the data so the countries with largest number of observations are at the top. Also, remove observations with no location.


#### Recap: Tidyverse vs Base R

To review, tidyverse is just another way to code in R with new functions and a framework based on the pipe `%>%`. You can always use a combination of both base R functions and tidyverse functions. 

There are some great resources on tidyverse functions. Here are a couple: 

- **[Beginner's tidyverse cheat sheet](https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf)**. The `dplyr` section includes the functions we focused on today. The `ggplot2` section is what we'll go over in the next section.
- **[Introduction to tidyverse](https://brianward1428.medium.com/introduction-to-tidyverse-7b3dbf2337d5)** includes a more comprehensive overview of tidyverse, from reading in data to applying essential functions (what we did today) to more advanced topics such as reshaping data frames.

