## Applying basic stats

Scholars will have a sense for how to implement more advanced statistical tests in R

```{r, include=FALSE}
library(tidyverse)
```

?

```{r}
head(msleep)
```

basic stats
describe quantitative variables

```{r}
median(msleep$brainwt, na.rm = TRUE)
mean(msleep$brainwt, na.rm = TRUE)
sd(msleep$brainwt, na.rm = TRUE)
var(msleep$brainwt, na.rm = TRUE)
```

?

```{r}
median(msleep$bodywt, na.rm = TRUE)
mean(msleep$bodywt, na.rm = TRUE)
sd(msleep$bodywt, na.rm = TRUE)
var(msleep$bodywt, na.rm = TRUE)
```

?

```{r}
hist(msleep$brainwt)
```

?

```{r}
cor(msleep$brainwt, msleep$bodywt)
```

know how to deal with NAs

```{r}
summary(msleep)
```

?

```{r}
msleep2 <- msleep %>% drop_na(brainwt)
dim(msleep)
dim(msleep2)
```

calculate confidence intervals

```{r}
#?
```

calculate t-test

```{r}
#?
```

calculate ANOVA

```{r}
#?
```

identify and use correlation functions
use different kinds of correlation

```{r}
cor(msleep2$brainwt, msleep2$bodywt)
```

```{r}
library(Hmisc)
rcorr(msleep2$brainwt, msleep2$bodywt, type = "pearson")
rcorr(msleep2$brainwt, msleep2$bodywt, type = "spearman")
```

identify and use linear regression lm() using both raw variable and formulaic interfaces

```{r}
my.mod <- lm(formula = brainwt ~ bodywt, data = msleep2)
summary(my.mod)
```

```{r}
my.mod2 <- lm(formula = sleep_total ~ bodywt + brainwt, data = msleep2)
summary(my.mod2)
```

```{r}
my.mod2 <- lm(formula = sleep_total ~ bodywt + brainwt, data = msleep2)
summary(my.mod2)
```
