## Exploring other plots

```{r,echo=FALSE,message=FALSE}
library(ggplot2)
library(tidyverse)
msleep2 <- msleep %>% mutate(brainwt_log = log(brainwt))
```

Let's consider how we make other plots besides a scatterplot.

Scatterplots are a great way to look at two quantitative (numerical) values at the same time to observe patterns (i.e. correlations) between the variables or to identify interesting outliers.

However, other plots may be more useful to look at differing numbers of variables (i.e. one quantitative variable) or different types of variables (i.e. qualitative or categorical data).

We can create histograms in ggplot2 that are more aesthetically pleasing than the default `hist` function. This shows the distribution of one quantitative variable.

```{r}
ggplot(msleep, aes(x = sleep_total)) +
    geom_histogram(bins = 10)
```

### Using categorical data

We can look at how many individuals in the dataset fall into each category, such as feeding behavior.

```{r}
ggplot(data = msleep, aes(x = vore)) + 
  geom_bar()
```

We can look at how many total sleep distribution is distributed within each category of feeding behavior using separate box plots.

```{r}
ggplot(data = msleep, aes(x = vore, y = sleep_total)) + 
  geom_boxplot()
```

### Visualizing between groups

By adding a different parameter to `fill` in the `aes` we define throgh the `ggplot` function, we can separate out histograms according to different groupings such as here where we look at how total sleep distribution is distributed according to feeding behavior.

```{r}
ggplot(msleep, aes(x = sleep_total, fill = vore)) +
    geom_histogram(bins = 10)
```

We can even drill down further and separate out another category, so that we can look at each feeding behavior and each conservation status (e.g. how the total sleep duration of domesticated herbivores compares to least concern carnivores).

```{r}
ggplot(msleep, aes(x = vore, y = sleep_total, color = conservation)) +
    geom_boxplot() 
```
