## Saving figures locally

```{r,echo=FALSE,message=FALSE}
library(ggplot2)
library(tidyverse)
msleep2 <- msleep %>% mutate(brainwt_log = log(brainwt))
```

If you want to save your images as a local individual file on your computer, there are two easy ways to do that.

The first method uses `ggsave` to save the most recent ggplot figure you generated.

```{r, eval=FALSE}
ggplot(msleep2, aes(x = brainwt_log, y = sleep_rem)) + geom_point() +
  geom_smooth(method = "lm")
ggsave("plot.png", width = 5, height = 5)
```

This function will save wherever your directory is currently. Check with `getwd()` and change with `setwd(folder name)`. You can also provide a precise file path in the new file name.

```{r, eval=FALSE}
ggsave("~/Downloads/plot.png", width = 5, height = 5)
```

Here is an alternative method for saving your figures:

```{r, eval=FALSE, warning=FALSE}
pdf("plot.pdf") # creates the file
# png() also works if you want a different file format
ggplot(msleep2, aes(x = brainwt_log, y = sleep_rem)) + geom_point() +
  geom_smooth(method = "lm")
dev.off() # finishes editing the file
```

Any changes to the figure that are contained between the initial creation of the figure (i.e. the `pdf` command) and the `dev.off` command will be included in the final saved image. However, the figure is being printed directly to the file it is writing and won't appear elsewhere.

