## Writing modular code

What is modularity of code?

Why?

think more clearly than “hardcoding” stuff

change things more easily without forgetting everything

lets you operate at a higher level, by tweaking parameters in one place

<!-- Exercise -->

### Calculating standard error

Let's calculate the standard error of a sample of values:

```{r,stderr_fail,error=T}
values <- c(4,3,2,2,5,3,6,2,2,4)
stderr(values)
```

Er ... what? What is `stderr` doing?

`?stderr`

Write out the maths for the standard error of the mean

calcuate it

```{r,stderr_calc}
sd(values)/sqrt(length(values)-1)
```

write a function that does it

what do we save it as?

test it

### More complex modules

More examples/exercise

show a simulation of something…. genetic drift of a virus replicating?

lineage G1312F

exercise - wrap the entire analysis as a function

talk about ease of calling

gotcha about print() the plots

exercise - make a parameter and argument

ease of tweaking this

exercise - break into subfunctions, generate and plot

ease of changing models
