## Packages are chunks of code (and sometimes data)

Let’s take a look at your code from the last few days. How would you share this with the other people in the course?

R, like other languages, is built on a package system - if you wrap up your code in particular expected ways and put it in particular expected places, it is very easy for others to get and use your code

side note link for interested students, easy of making pakcages (for advanced advanced)

### first stop - CRAN

[CRAN]()

```{r,eval=F}
library("stringdist")
```

exercise - get the stringdist library, use it

exercise - get a new viz layer, use it


### How does this package work?

?

??

etc


### next stop - github

```{r,eval=F}
library("devtools")
```

or `remotes`?

exercise - pull a package off github... ? maybe some sort of viz again?

Exercise - install a package from github and learn how to use it

- [Ghibli color palette](https://github.com/ewenme/ghibli)
- [Sports field plotting](https://github.com/rossdrucker/sportyR)
- [Ridgeplots in ggplot2](https://github.com/wilkelab/ggridges)


<!-- remotes::install_github("wilkelab/ggridges" -->

problems with this - security, eh

### finally - Bioconductor 😬

[How to install Bioconductor]()

Scholars will know how to install Bioconductor package from CRAN

Scholars will know how to find and install packages from Bioconductor

explore the namespace for a package, ie ?packagename:: and TAB or ?packagename::

how to print out all the packages you have loaded


## Finding more help and documentation

Need a problem

```{r,eval=F}
?geom_line
```

```{r,eval=F}
?stringdist
```

Objective X.X: Scholars will know how to get official docs ( ? ) and how to skim/read these

Objective X.X: Scholars will see how easy it is to use a search engine to try and find other answers

Objective X.X: Scholars will understand the intention of the Slack being a learning community, or something (set expectations about this)

Source up problems from the class?

Demo a weird plot idea, search for how to do it on stack overflow?

Maybe how to change the color of facet labels in a ggplot facet graph, that's a bit tricky and demonstrates why base R is important…

## Reproducibility and sharing

think about this

think about the next person coming along, or yourself in 3 months

links on syllabus for more info

all sorts of automated stuff you can do



