## Packages - install and use

Let’s take a look at your code from the last few days. 
How would you share this with the other people in the course?

R, like other languages, is built on a package system - if you wrap up your 
code in particular expected ways and put it in particular expected places, 
it is very easy for others to get and use your code

Packages are 
[said to be easy to make](https://ourcodingclub.github.io/tutorials/writing-r-package/),
[from scratch](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/).
There's more [comprehensive instruction here](https://r-pkgs.org/).
Personally, the author has never had cause to make one, but I can report that
they're very easy to use!

---

### CRAN is the canonical R package resource

The Comprehensive R Archive Netowrk 
[CRAN](https://cran.r-project.org/web/packages/)
is an integral part of the R environment.
This is a global and free resource for storing and distributing code
as R packages - well documented and designed chunks of code.

You can install packages using the RStudio IDE, but it's most important
to know how to do it in R - what's happening just under the hood.

Use `library` to load packages. 
You should be able to load a package you've
already installed. 
Call `library()` function, with an argument of the library name,
as a string.

Try loading this one:

```{r,eval=F}
library("stringdist")
library(stringdist)
```

Don't have it? You can look for packages by using a search engine, or finding
suggestions in the documentation, papers. Once you have a package name, use

```{r,eval=F}
install.packages("stringdist")
```

to install it. You may need to select a "mirror". This is just where you
are downloading from, usually just go with "0-Cloud"
^[I have no idea where this is, never heard of that country!].

What did this do? 
It retrieves and downloads the package's code and documentation to
what's called your library, 


- get the stringdist library, use it
- get a new viz layer, use it

---

### Using packages without `library()`-ing them

Some packages you don't really want to have to go through the trouble of
loading, you'll just want one function. 
For example, this website is generated with one function, `render_book()`
in the package `bookdown`. Instead of:

```{r,eval=F}
library(bookdown)
render_book()
```

I can save a line by just using `::`

```{r,eval=F}
bookdown::render_book()
```

This means I am looking in library `bookdown` for function `render_book`.
The function is hidden again right afterwards, but it is useful if you just
want to run one function from a library, once.


---

### Github is a common place for sharing packges in development

Installing from github requires some different programming -
and there's libraries with functions for this.

```{r,eval=F}
library("remotes")
```

This used to be part of the `devtools` package, but the authors have been
"uncoupling" these to make each package simple and composable.

#### Let's install a package off of github.

Here's a cool [Ghibli color palette](https://github.com/ewenme/ghibli),
so you can change sets of colors for your plots.
Try installing it with:

```{r,eval=F}
remotes::install_github("ewenme/ghibli")
```

Then adapt one of your previous plots (that has colors) to use these new
color palettes.


Exercise - install a package from github and learn how to use it

- [Sports field plotting](https://github.com/rossdrucker/sportyR)
- [Ridgeplots in ggplot2](https://github.com/wilkelab/ggridges)


<!-- remotes::install_github("wilkelab/ggridges" -->

problems with this - security, eh

---

### Bioconductor is a great place for bio-related packages 😬 

Bioconductor is a giant repository of just scientific, well bio-focused,
packages. It's where a lot of bioinformatics software gets published.
However, it has it's own package system and thus package manager.
You will have to install it ... from CRAN. Like so:

```{r,eval=F}
install.packages("BiocManager")
```

That installs a package with a function that then installs Bioconductor
packages. It's actually easy. Let's give it a try by installing and
using the `limma` package to analyze some gene expression data.

If you're recall from before, we can use `::` to avoid having to actually
`library` the whole package:

```{r,eval=F}
BiocManager::install()
```



[How to install Bioconductor]()

Scholars will know how to install Bioconductor package from CRAN

Scholars will know how to find and install packages from Bioconductor

---

As demo, install DESeq2 and start to work through a tutorial with one of
their datasets.

---

### What packages do you have already? 

explore the namespace for a package, ie ?packagename:: and TAB or ?packagename::

how to print out all the packages you have loaded

---

### Finding more help and documentation

Need a problem

```{r,eval=F}
?geom_line
```

```{r,eval=F}
?stringdist
```

Objective X.X: Scholars will know how to get official docs ( ? ) and how to skim/read these

Objective X.X: Scholars will see how easy it is to use a search engine to try and find other answers

Objective X.X: Scholars will understand the intention of the Slack being a learning community, or something (set expectations about this)

Source up problems from the class?

Demo a weird plot idea, search for how to do it on stack overflow?

Maybe how to change the color of facet labels in a ggplot facet graph, that's a bit tricky and demonstrates why base R is important…

---
