
<!--

```{=html}
<div class="incremental">
```
```{=html}
</div>
```

-->

## Functions redux - why they're so great

Functions are handy.
For example, let's calculate the standard error of a sample of values:

```{=html}
<div class="incremental">
```
```{r,stderr_fail,error=T}
values <- c(4,3,2,2,5,3,6,2,2,4)
stderr(values)
```
```{=html}
</div>
```

```{=html}
<div class="incremental">
```

Er ... what? What does `stderr` do?

`?stderr`

```{=html}
</div>
```

---

How do you calculate the standard error of the mean of a sample?

```{=html}
<div class="incremental">
```
```{r,stderr_calc,cache=T}
sd(values)/sqrt(length(values))
```

Do we want to type this each time, copy and paste the code each time?

```{r,stderr_calc_2,cache=T}
values2 <- c(10,30,20)
sd(values2)/sqrt(length(values))
```

This can lead to errors in writing, small incorrect parts.
It also leaves all the complexity to the reader, so it makes it harder to
understand when reading code - hard to notice small details.

Where do we save it, how do we share it?

Copy and pasting from a file is okay, but it takes up a lot more work/space.
We could email around a file of code chunks, or share on github.

How do we make sure it works?

This is hard to do with copy-paste code chunks. Changing the chunk to work
on a new input can be an opportunity for introducing errors, typos.

```{=html}
</div>
```

---

#### Write it as a function!

What are the parameters? Are any defaults set?
What does it return?

These can be copy and pasted, or `source()`'d from a `.R` file.
Once you run it, you can test it and change inputs easily.
Later, you could change the function and keep the same script of using the
function.

```{=html}
<div class="incremental">
```
```{r,stderr_test,cache=T}
sez <- function(x) {sd(x)/sqrt(length(x))}
sez(values)

sez(rnorm(10))
sez(rnorm(100))
sez(rpois(1e3,3))
```
```{=html}
</div>
```

---

## Using functions to write modular code

How might you carry out the workflow from the previous work to
calculate protein sequence lengths?

```{=html}
<div class="incremental">
```

One way is to flatten out all the tasks, 
and to script each individual task each time.
This requires the author, user, and reader to understand a lot of complexity.

```{=html}
<div class="mermaid">
graph TD
    classDef default line-height:10px;
    classDef three line-height:10px,fill:yellow;
    you[analyst] --> B[finding files] &    C[making a vector] &    D[looping] &    E[reading files] &    F[getting protein seq] &    G[calculting and saving nchar] &    H[plotting nchar]
    B:::three;
    C:::three;
    D:::three;
    E:::three;
    F:::three;
    G:::three;
    H:::three;
</div>
```

Another way to approach this is to cluster them into a hierarchy of modules.

```{=html}
<div class="mermaid">
graph TD
    classDef default line-height:10px;
    classDef one  line-height:10px,fill:pink;
    classDef two line-height:10px,fill:cyan;
    classDef three line-height:10px,fill:yellow;
    you2[analyst] --> find & read & plot;
    find:::one --> B;
    B[finding files]:::three;
    read:::one --> looping & reading;
    looping:::two --> C & D;
    C[making a vector]:::three;
    D[looping]:::three;
    reading[reading and calc]:::two --> E & F & G;
    E[reading files]:::three;
    F[getting protein seq]:::three;
    G[calculting and saving nchar]:::three;
    plot:::one --> H;
    H[plotting nchar]:::three;
</div>
```

In this organization, the analyst can operate at levels of 
<span style="color:magenta">steps</span>,
<span style="color:cyan">modules</span>, and
<span style="color:gold">specific instructions</span>,
depending on what is needed.

```{=html}
</div>
```

---

Organizing your workflows into composable modules lets you extend these to
un-ancipiated complexity. You could imaging using these steps or modules:

```{=html}
<div class="mermaid">
graph LR
    classDef default line-height:12px;
    vp[read viral proteins] --> cl[calculate lengths] --> hist[histogram]
</div>
```

in new ways by composing the elements together, to analyze a different source
of proteins, with a new analysis, with similar plots:

```{=html}
<div class="mermaid">
graph LR
    classDef default line-height:12px;
    vp[read viral proteins] --> cl[calculate lengths] --> hist[histogram]
    hp[read human proteins] --> cl --> boxplot[boxplot]
    vp --> cgc[calculate GC%] --> gchist[histogram]
    hp --> cgc --> gcboxplot[boxplot]
</div>
```

---

To do this, you want to:

- try to not "hardcode" things - if it's a number, consider if it can be a
    *parameter* that is "passed in" as an argument
- write things as functions
- think of standard ways that functions can accept and return values

Consider, the `list.files()` and `hist()` functions are already built this way!
How about a `ggplot2` style boxplot?

Lets code that up inside a function. What should the function take,
what should it do?


### More complex modules

More examples/exercise

show a simulation of something…. genetic drift of a virus replicating?

lineage G1312F

exercise - wrap the entire analysis as a function

talk about ease of calling

gotcha about print() the plots

exercise - make a parameter and argument

ease of tweaking this

exercise - break into subfunctions, generate and plot

ease of changing models



Apply is a thing

Scholars will be familiar with that folks have strong opinions about using apply instead of for loops

Scholars will understand that getting the code done is the most important thing, but that good use of apply can be something to develop later to make code run faster, cleaner

Scholars will be able to read what an example of apply or s/lapply is doing, so that they can read examples online

Scholars will have heard the term "lambda function" ?

Scholars will understand that functions can be one-off (ie lambda, function(x){} ), and that's okay


<script src="https://cdn.jsdelivr.net/npm/mermaid/dist/mermaid.min.js"></script>
<script>
mermaid.initialize({
    startOnLoad: false, // don't render initially, instead....
    logLevel: 3, // for debugging
});
function initMermaid(s) { 
  var diagrams = document.querySelectorAll('.mermaid');
  var i;
  for(i=0;i<diagrams.length;i++){
    if(diagrams[i].offsetWidth>0){
      mermaid.init(undefined, diagrams[i]);
    }
  }
}
initMermaid();
</script>
