
## Writing (re)usable code

- Crack open your code - can you use it again?

- Can you adapt it to modify your question, feed in new data,
and modify the scientifically-important bits easily?

---

#### Examples

Let's look at two chunks of code from 
[a paper](https://github.com/sashaflevy/PPiSeq) (lightly edited).
The experiment is counting barcoded lineages of yeast cells to estimate
"PPIs" (protein-protein interactions)
^[Ask Darach if you want details, or [read the paper](doi.org/10.7554/eLife.62365)]

Here's an example of one style of writing R script:

```{r,example1,eval=F}
# filter out bad barcode lineages ( <= 2 time points counts > 0, or maximum of each time point <= 5 or total counts of a lineage < 10)
bad_index = rep(0, nrow(DBC_known_counts))
for(i in 1:length(bad_index)){
        counts = as.numeric(DBC_known_counts[i, 4:8])
        if (length(which(counts != 0)) < 3 | max(counts) < 5 | sum(counts) < 10){
                bad_index[i] = 1
        }
}
length(which(bad_index == 1)) # 1447775
```

<!-- exercise -->

- What is going on here?
- How do you feed in new data?
- How do you run this multiple times?
- How do you change the logic?

---

Here's another example from the same author:

```{r,example2,eval=F}
H202_Output <- PPI_calling_sigmoid(
    PPI_multiple=dataFrameReader_T("/Volumes/zmliu_02/PPiseq_03/H2O2/counts/PPI_multiple_p.values.csv"),
    specific_PPV=c(seq(0.5,0.58,by=0.02),seq(0.6,0.8,by=0.01),seq(0.82,0.9,by=0.02)),
    Fitness=fitness(0,1,0.01),
    p_value=seq(-4,0,0.1),
    Neg_number_PPI=6e4,
    Neg_ref_number=50, 
    p_threshold=-4
    p_loc=6, 
)
```

<!-- exercise -->

Same questions.

---

What are some differences?

What's useful for you?

What's a lot of effort to do?

---

### Style guides can be inspiring

- [tidyverse style guide](https://style.tidyverse.org/)
- [google-specific changes](https://google.github.io/styleguide/Rguide.html)
- [Jean Fan's](https://jef.works/R-style-guide/)
- search for "R style guide"

What style should you use?

Be inconsistently consistent!
Balance for yourself:

- How easy is it to write?
- How easy is it for you to read?
- How easy is it for others to read?
- How similar is it to what everyone else is doing (a very good thing)?

But most importantly, use what folks around you are using.
Be lazy, imitate others!

---

<!-- exercise -->

- How would you describe your code writing style?
- How do you name things?
- How good are your comments?
- What ideas would you like to incorporate?

---
