## Packages are chunks of code (and sometimes data)

Let’s take a look at your code from the last few days. 
How would you share this with the other people in the course?

R, like other languages, is built on a package system - if you wrap up your 
code in particular expected ways and put it in particular expected places, 
it is very easy for others to get and use your code

Packages are 
[said to be easy to make](https://ourcodingclub.github.io/tutorials/writing-r-package/),
but they're very easy to use.

---

### CRAN is the canonical R package resource

[CRAN]()

historical context

```{r,eval=F}
library("stringdist")
```

Call `library()` function, with an argument of the library name,
as a string.

- get the stringdist library, use it
- get a new viz layer, use it

---

### github is a common place for sharing code in development

Installing from github requires some different programming -
and there's libraries with functions for this.

```{r,eval=F}
library("devtools")
```

or `remotes`?

exercise - pull a package off github... ? maybe some sort of viz again?

Exercise - install a package from github and learn how to use it

- [Ghibli color palette](https://github.com/ewenme/ghibli)
- [Sports field plotting](https://github.com/rossdrucker/sportyR)
- [Ridgeplots in ggplot2](https://github.com/wilkelab/ggridges)


<!-- remotes::install_github("wilkelab/ggridges" -->

problems with this - security, eh

### finally - Bioconductor 😬

[How to install Bioconductor]()

Scholars will know how to install Bioconductor package from CRAN

Scholars will know how to find and install packages from Bioconductor

explore the namespace for a package, ie ?packagename:: and TAB or ?packagename::

how to print out all the packages you have loaded


## Finding more help and documentation

Need a problem

```{r,eval=F}
?geom_line
```

```{r,eval=F}
?stringdist
```

Objective X.X: Scholars will know how to get official docs ( ? ) and how to skim/read these

Objective X.X: Scholars will see how easy it is to use a search engine to try and find other answers

Objective X.X: Scholars will understand the intention of the Slack being a learning community, or something (set expectations about this)

Source up problems from the class?

Demo a weird plot idea, search for how to do it on stack overflow?

Maybe how to change the color of facet labels in a ggplot facet graph, that's a bit tricky and demonstrates why base R is important…

## Reproducibility and sharing

think about this

think about the next person coming along, or yourself in 3 months

links on syllabus for more info

all sorts of automated stuff you can do



