## Reproducibile sharing - more `rmarkdown` and 

How are you saving your notes for this class?
<!-- rmd you better be... -->

How do you usually save notes/code/instructions?

How do you share this with others?

### rmarkdown package {.tabset}

#### Background

The `rmarkdown` package was created by
[Yihui Xie](https://www.rstudio.com/speakers/yihui-xie/),
extending from previous work and ideas in 
`knitr` and `sweave`.

It's a way of mixing writing with code, such that
you can run the code and it makes a pretty doc
(and [more](https://rmarkdown.rstudio.com/gallery.html)).

video1 overview
<!-- 
motivation / ie what it do
video1 background / broad context
yihui xie - grad student, wrote a bunch of packages, now works at Rstudio (pattern here…)
sweave -> knitr -> rmarkdown
how I’ve used it, how you could use it
-->

#### Chunk options

You can set chunk-level options in an Rmd, for
each code chunk. Such as:

````
```{r, name_of_chunk, cache=T, fig.width=3, fig.height=2, error=F, warning=F, fig.align="right"}`r ''`
x <- 1:10
y <- 1:10
plot(x,y)
```
````


```{r, name_of_chunk, cache=T, fig.width=3,fig.height=2, error=F, warning=F,  fig.align="right"}
x <- 1:10
y <- 1:10
plot(x,y)
```

It is possible to enable figure captions.

video2 chunk options
<!--
how to cache=T
how to hide warnings/errors
how to change figure sizes
fig.align=”right”
-->

#### Document options

You can set document-level options to. To do this,
you create what's called a 
[YAML](yaml.org) header, like so:

````
---
title: "Titled"
author: "yours"
---
````

You need three hyphens to open and close it.
Put it at the very beginning. You can define
quite a few options, including *themes*,
like so:


````
---
title: "Titled"
author: "yours"

---
````

video3 head options 
<!--title
authors
date, Sys.date
format …. slides !
*** themes **** &&%
-->

#### How it works, sort of... 

video4 how it does what it’s doing
<!--what it’s doing behind the scenes - a diagram ( on paper )
code gets pulled out and run, top to bottom
-->

`rmarkdown` parses an Rmd file to extract out
the R code. It runs this, using the chunk and
other options to control this, then makes outputs
from different chunks (text and plots).
It then sticks the text and images into a
`markdown` document, and uses a program called
`pandoc` (included inside Rstudio) to turn that
into HTML, PDF, and/or slides.

### {-}

### Sharing with others

What does someone need to re-run your analyses?

<!-- R, Rstudio, packages, scripts -->

- R (often Rstudio)
- what packages you're using
- 

```{r,sessioninfo,cache=T}
devtools::session_info()
```

why the hell we want others to use our code?
Well I'll tell you why - the general intellect

think about this

think about the next person coming along, or yourself in 3 months

links on syllabus for more info

all sorts of automated stuff you can do




