---
title: "Worksheet for Day 4 - tidyverse and visualizations"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Instructions for this Worksheet

Please be sure to write in your code or response underneath each prompt. Be sure that the code is able to execute completely and that the result you expect is shown in the resulting html file.

------

#### Section 3.1: Reading and processing data

##### Section 3.1.1: Loading a file into R

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.1.2: Handling issues in data

**Task:** ?

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.2: Using tidyverse

------

#### Section 3.3: Using tidyverse

------

#### Synchronous Activities

**If you reach this point during the asynchronous session, please stop and wait until the synchronous session starts.**

**Task:** ???

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.4: Making plots with ggplot2

##### Section 3.4.1: Getting started with a ggplot

**Task:** Starting from the `msleep` dataset, pick two quantitative variables with which to create a scatterplot. Save this plot as a variable and then display this figure. Describe any patterns or outliers you see in this plot.

```{r}
library(ggplot2)
# Your code here
```

*Your response here*

**Task:** Starting from the `msleep` dataset, pick at least one quantitative variable to scale or transform. Make this new vector into a column in your dataset named `msleep2`. You may need to use the `mutate` function.

```{r}
# Your code here
```

**Task:** Starting from this `msleep2` dataset, use your new column and one other quantitative variable to create a scatterplot. Save this plot as a variable and then display this figure. Describe any patterns or outliers you see in this plot.

```{r}
# Your code here
```

*Your response here*


##### Section 3.4.2: Changing plot aesthetics

**Task:** Working from the code you used to generate one of these scatterplots, create a new scatterplot where you uniformly change the shape, color, and/or size of the points.

```{r}
# Your code here
```

**Task:** Working from the code you used to generate one of these scatterplots, change the labels on this plot.

```{r}
# Your code here
```

**Task:** Working from the code you used to generate one of these scatterplots, create a new scatterplot where you have the color dictated by a quantitative variable in the `msleep` or `msleep2` dataset. Describe any patterns or outliers you see in this plot.

```{r}
# Your code here
```

*Your response here*


##### Section 3.4.3: Exploring simple plots

**Task:** Create a new histogram looking at a different quantitative variable either in the `msleep` dataset or the `iris` dataset. Describe the distribution that you see: what is the range of values? Is there one peak or more than one peak? What appears to be the center/average?

```{r}
# Your code here
```

*Your response here*


**Task:** For your histogram, try changing the value of `bins` in the `geom_histogram` function. How does this change your plot and how you might interpret the distribution?

```{r}
# Your code here
```

*Your response here*

**Task:** Create a new barplot looking at the counts of a categorical variable within either the `msleep` or `iris` dataset. Describe the numbers of samples that you see.

```{r}
# Your code here
```

*Your response here*

##### Section 3.4.4: Visualizing between groups

**Task:** Create a different boxplot for either the `iris` dataset or the `msleep` dataset, by using either a different categorical variable or a different quantiative variable (or both!). Describe what differences you see, if any, between the groups.

```{r}
# Your code here
```

*Your response here*

**Task:** Create an analogous histogram with faceting to display the differences between the same groups along the same quantiative variable, just using a different kind of plot. Which plot would you use between these two, and why?

```{r}
# Your code here
```

*Your response here*


##### Section 3.4.5: Generating heatmaps

**Task:** Create a new tabular dataset from the `iris` dataset where you extract just two of the quantatitive variables. Save this new dataset to a variable.

```{r}
# Your code here
```

*Your response here*

**Task:** Using this new dataset with just two columns, create a basic heatmap. Remember to provide the names of the columns using the `labCol` argument for the `heatmap` function. What patterns/outliers do you observe, if any, in this heatmap?

```{r}
# Your code here
```

*Your response here*

**Task:** For this same heatmap, change the `scale` parameter in the `heatmap` function to either `"row"` or `"col"` and observe the differences. What do you notice? You may want to look into the function using `?heatmap` to check what the argument does.

```{r}
# Your code here
```

*Your response here*

**Task:** For this same heatmap, add the `Species` data from the original `iris` dataset to the heatmap using the `labRow` argument in this function. What patterns if any do you now notice?

```{r}
# Your code here
```

*Your response here*


**Task:** For a challenge, you can try adding more quantitative variables from `iris` to your numeric dataset and replotting the heatmap. Share any observations on how this changes your final heatmap.

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.5: Making scientific figures

##### Section 3.5.1: Plotting error bars

**Task:** Choose either a different categorical variable or a different quantitative variable in the `msleep` dataset with which to generate a new summary dataframe. This dataframe will contain the average, lower bound, and upper bound measurement within each group. You will want to use the `group_by`, `summarize`, and `mean_se` functions.

```{r}
# Your code here
```

**Task:** With your new summary dataframe, create a barplot that shows the average measurement in each categorical group. What differences do you see if any between the groups?

```{r}
# Your code here
```

*Your response here*

**Task:** For the barplot you just generated, add the errorbars. How does this change your interpretation in the last task, if at all?

```{r}
# Your code here
```

*Your response here*

**Task:** For a challenge, repeat these last few steps with the `iris` dataset to show differences in any quantitative variable between the `Species` in this data. Describe your findings.

```{r}
# Your code here
```

*Your response here*


##### Section 3.5.2: Showing trends in data

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.5.3: Saving figures locally

**Task:** ?

```{r}
# Your code here
```

*Your response here*

------

#### Synchronous Activities

**If you reach this point during the asynchronous session, please stop and wait until the synchronous session starts.**

**Task:** ???

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.6: Applying basic stats

##### Section 3.6.1: Describing quantitative variables

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.6.2: Finding differences in data

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.6.3: Identifying correlations

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.6.4: Producing linear models

**Task:** ?

```{r}
# Your code here
```

*Your response here*
