---
title: "Worksheet for Day 5 - building workflows"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This worksheet was created by [darachm](https://github.com/darachm) and [Samson](https://github.com/samson920) with a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 License](https://github.com/darachm/dll-r/blob/main/LICENSE.md).

#### Instructions for this Worksheet

Write your code and/or your response underneath each task prompt. To render your worksheet, click on the Knit button in Rstudio. Be sure that your code is able to execute completely and that the result you expect is shown in the resulting html file.

------

#### Section 4.2: Loops to read many files

##### Section 4.2.1: Reading in one file

**Task A:** ???

```{r}
# Your code here
```

*Your response here*

#### End of Class Summary

Now is an opportunity for you to put together everything that you've learned to start asking questions about data. There are a few datasets listed below, along with questions about the data. First, design some analyses you can do to answer the question. Some examples might be comparing the means between certain populations, plotting the distribution of a variable (or multiple variables), or . For all of these datasets, make sure to explore the data a little bit with commands like head() and tail() before jumping into your analysis so you're familiar with the data.

**Iris Dataset**

You can load this dataset with `data(iris)`

What variable is most useful for distinguishing between species?

What species has the biggest variability in flower size?



**Alzheimer's Microarray Analysis**

These data are originally from https://www.kaggle.com/andrewgao/alzheimer-microarray-analysis. You can download them from https://github.com/darachm/dll-r/tree/main/data

What gene is most associated with Alzheimer's disease (based on the unadjusted p-value)?

How many genes have unadjusted p-values less than 0.05?

Do some reading on p-values. Why do you think the adjusted p-value is included in the data?


**Mortality Data**

The CDC tracks data on mortality in the United States. We have downloaded a dataset on mortality by state, by week across a few years. You can download the data from https://github.com/darachm/dll-r/tree/main/data, though the data were originally from https://www.kaggle.com/codebreaker619/usa-weekly-counts-of-deaths-by-state-and-causes.

What state has the highest number of deaths per year based on the data reported here? Do you think this is a fair way to look at the data? What can you do to help make the analysis more fair across states if you wanted to understand which states had the highest death rates?

What cause of death has increased the most from 2014 to 2018? Which has decreased?

What week (based on the MMWR week column) generally has the highest mortality?
