---
title: "Worksheet for Day 4 - tidyverse and visualizations"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Instructions for this Worksheet

Please be sure to write in your code or response underneath each prompt. Be sure that the code is able to execute completely and that the result you expect is shown in the resulting html file.

------

#### Section 3.1: Reading and processing data

##### Section 3.1.1: Loading a file into R

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.1.2: Handling issues in data

**Task:** ?

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.2: Using tidyverse

------

#### Section 3.3: Using tidyverse

------

#### Synchronous Activities

**If you reach this point during the asynchronous session, please stop and wait until the synchronous session starts.**

**Task:** ???

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.4: Making plots with ggplot2

##### Section 3.4.1: Getting started with a ggplot

**Task:** Starting from the `msleep` dataset, pick two quantitative variables with which to create a scatterplot. Save this plot as a variable and then display this figure. Describe any patterns or outliers you see in this plot.

```{r}
library(ggplot2)
# Your code here
```

*Your response here*

**Task:** Starting from the `msleep` dataset, pick at least one quantitative variable to scale or transform. Make this new vector into a column in your dataset named `msleep2`. You may need to use the `mutate` function.

```{r}
# Your code here
```

**Task:** Starting from this `msleep2` dataset, use your new column and one other quantitative variable to create a scatterplot. Save this plot as a variable and then display this figure. Describe any patterns or outliers you see in this plot.

```{r}
# Your code here
```

*Your response here*


##### Section 3.4.2: Changing plot aesthetics

**Task:** Working from the code you used to generate one of these scatterplots, create a new scatterplot where you uniformly change the shape, color, and/or size of the points.

```{r}
# Your code here
```

**Task:** Working from the code you used to generate one of these scatterplots, change the labels on this plot.

```{r}
# Your code here
```

**Task:** Working from the code you used to generate one of these scatterplots, create a new scatterplot where you have the color dictated by a quantitative variable in the `msleep` or `msleep2` dataset. Describe any patterns or outliers you see in this plot.

```{r}
# Your code here
```

*Your response here*


##### Section 3.4.3: Exploring simple plots

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.4.4: Visualizing between groups

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.4.5: Generating heatmaps

**Task:** ?

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.5: Making scientific figures

##### Section 3.5.1: Plotting error bars

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.5.2: Showing trends in data

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.5.3: Saving figures locally

**Task:** ?

```{r}
# Your code here
```

*Your response here*

------

#### Synchronous Activities

**If you reach this point during the asynchronous session, please stop and wait until the synchronous session starts.**

**Task:** ???

```{r}
# Your code here
```

*Your response here*

------

#### Section 3.6: Applying basic stats

##### Section 3.6.1: Describing quantitative variables

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.6.2: Finding differences in data

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.6.3: Identifying correlations

**Task:** ?

```{r}
# Your code here
```

*Your response here*

##### Section 3.6.4: Producing linear models

**Task:** ?

```{r}
# Your code here
```

*Your response here*
