---
title: "Worksheet for Day 3 - introduction to R"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Instructions for this Worksheet

Please be sure to write in your code or response underneath each prompt. Be sure that the code is able to execute completely and that the result you expect is shown in the resulting html file.

------

#### Section 2.1: Navigating Rstudio

------

#### Section 2.2: Using Rmds

------

#### Section 2.3: Getting started in R

##### Section 2.3.1: Getting help in R

**Task:** Use the `help` function to look up another function that you have seen used before, or Google popular functions in R. The `help` function will cause documentation to appear in a side panel of RStudio when it is run within the Rmd code chunk. Which function did you look up? What arguments does it take and what does it output?

```{r}
?rm
help(rm)

# Your code here
```

*Your response here*

##### Section 2.3.2: Using variables and datatypes

**Task:** Try assigning multiple other variables below. See what types of data you can save and what variable names are acceptable in R. Note when you get a warning or error message.

```{r}
save.string <- "hello"
save.bool <- TRUE

# Your code here
```

*Your response here*

**Task:** Check that you can only remove one or two of the new variables you created, but that you can see the other ones still retained when you call the `ls` function.

```{r}
# Your code here
```


##### Section 2.3.3: Using functions

**Task:** Pick one of the following functions to look into: `which`, `sum`, `sort`, `paste`, or `sample`. Research this function to identify its arguments and outputs. We recommend looking at the sections for Usage, Arguments, and Examples.

```{r}
# Your code here
```

*Your response here*

**Task:** For the function you picked, practice successfully using this function below without errors.

```{r}
# Your code here
```

*Your response here*

**Task:** For a challenge, look into a more complicated function to use such as one of the following: `write.csv`, `apply`, `set.seed`, or `detach`.

```{r}
# Your code here
```

*Your response here*

------

#### Synchronous Activities

**If you reach this point during the asynchronous session, please stop and wait until the synchronous session starts.**

**Task:** ???

```{r}
# Your code here
```

*Your response here*


------

#### Section 2.4: Working with vectors

##### Section 2.4.1: Building vectors

**Task:** Try making a vector that contains elements of different data types (character strings like "this" or numeric values like 12). Describe what happens in each of these cases.

```{r}
# Your code here
```

*Your response here*

**Task:** Make several vectors of varying number of elements and save them to variables. Practice combining these vectors together using the `c` function.

```{r}
# Your code here
```

**Task:** Make a vector that contains only odd numbers between 9 and 35. You may need to check what the `:` operator, `seq`, or `rep` functions do.

```{r}
# Your code here
```

**Task:** Make a vector that contains five of each number between 3 and 6. You may need to check what the `:` operator, `seq`, or `rep` functions do.

```{r}
# Your code here
```


##### Section 2.4.2: Indexing and subsetting vectors

**Task:** Remove the last element in the vector `all.heart.rates` and save the result in a new variable.

```{r}
heart.rates <- c(78, 68, 95, 82, 69, 63, 86, 74, 64, 62)
more.heart.rates <- c(86, 79, 64, 74, 80)
all.heart.rates <- c(heart.rates, more.heart.rates)
# Your code here
```

**Task:** Index into the vector `all.heart.rates` and access only the last three elements contained in that vector.

```{r}
# Your code here
```

**Task:** Index into the vector `all.heart.rates` and access only every other number contained in that vector. You may need to check what the `:` operator, `seq`, or `rep` functions do.

```{r}
# Your code here
```

##### Section 2.4.3: Using logic and logicals

**Task:** Describe what happens when you apply the `sum` function to the vector `check.flags` used below. Based on what the `sum` function does, what do you think is happening? 

```{r}
check.flags <- c(TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, TRUE)
# Your code here
```

*Your response here*

**Task:** Check whether any elements in the vector `all.heart.rates` are greater than 90.

```{r}
# Your code here
```

**Task:** Calculate how many elements in the `all.heart.rates` vector are less than 70.

```{r}
# Your code here
```

**Task:** Index into the vector `all.heart.rates` and access only the elements that are less than 70.

```{r}
# Your code here
```

**Task:** Index into the vector `all.heart.rates` and access only the elements that are less than the average heart rate in this sample. You may want to check what the `mean` function.

```{r}
# Your code here
```

**Task:** For a challenge, index into the vector `all.heart.rates` and access only the elements that are less than 85 **or** greater than 90.

```{r}
# Your code here
```

##### Section 2.4.4: Modifying vectors

**Task:** Figure out what happens when you try to add two vectors containing numeric values that are of different lengths (i.e. one vector contains three numbers, the other vector contains five numbers).

```{r}
# Your code here
```

**Task:** Take the vector `all.heart.rates`, which expresses beats per minute (bpm), and create a new vector that expresses these numbers as beats per second.

```{r}
# Your code here
```

**Task:** Assume that the vector `all.heart.rates` was collected with a device that only records about 90% of the beats. Multiply this vector to correct for this.

```{r}
# Your code here
```

**Task:** Your new calculation of `all.heart.rates` likely does not have integer values. Calculate either the rounded or significant figure values to get rid of the fractions.

```{r}
# Your code here
```

------

#### Section 2.5: Working with dataframes

##### Section 2.5.1: Indexing and subsetting dataframes

**Task:** Access into the `iris` dataframe and only pull out the first two rows of observations.

```{r}
# Your code here
```

**Task:** Access into the `iris` dataframe and only pull out the first column.

```{r}
# Your code here
```

**Task:** Create a new dataframe called `iris2` that has the last two columns of the original dataframe removed. 

```{r}
# Your code here
```

**Task:** Create a new dataframe called `iris3` that has the last row of the original dataframe removed. 

```{r}
# Your code here
```

**Task:** Based on the code provided below, access into the `iris` dataframe and only pull out the columns that refer to petal measurements.

```{r}
iris[, "Sepal.Length"]
# Your code here
```


##### Section 2.5.2: Exploring dataframes

**Task:** Investigate the number of rows and columns in the `msleep` dataset that comes with the `ggplot2` library. Describe your findings.

```{r}
library(ggplot2)
# Your code here
```

*Your response here*

**Task:** Investigate what the different columns are in the `msleep` dataset and what kinds of data they contain (i.e. which datatypes). Describe your findings.

```{r}
# Your code here
```

*Your response here*

**Task:** Investigate the range of quantitative variables in the `msleep` dataset. Describe your findings.

```{r}
# Your code here
```

*Your response here*


**Task:** Investigate the counts of categorical variables in the `msleep` dataset. Describe your findings.

```{r}
# Your code here
```

*Your response here*


##### Section 2.5.3: Building and modifying dataframes

**Task:** Working off of the code below and the `msleep.small` dataframe, try to add a new column to this dataset. Feel free to make up the data without worrying about it being factual!

```{r}
msleep.small <- msleep[1:7, c("name", "vore", "sleep_total", "brainwt")]
# Your code here
```

**Task:** Starting with the `msleep.small` dataframe, try to add a new row to this dataset. Feel free to make up the data without worrying about it being factual!

```{r}
# Your code here
```


**Task:** Adapting the code above used to make the `msleep.small` dataframe, try to add a new dataframe that pulls different rows than the first 7 in `msleep`. Save this dataframe as a different variable.

```{r}
# Your code here
```

**Task:** Working from the last task, merge the `msleep.small` dataframe and your new dataframe together.

```{r}
# Your code here
```


##### Section 2.5.4: Handling datatypes in dataframes

**Task:** Check the datatype of the `vore` column in the `msleep.small` dataframe that contains the feeding behavior of the mammals.

```{r}
# Your code here
```

*Your response here*

**Task:** Convert the `vore` column in the `msleep.small` dataframe to a different datatype. Describe anything that may have changed.

```{r}
# Your code here
```

*Your response here*

**Task:** Convert the `sleep_total` column in the `msleep.small` dataframe to a different datatype. Describe anything that may have changed.

```{r}
# Your code here
```

*Your response here*

------

#### Synchronous Activities

**If you reach this point during the asynchronous session, please stop and wait until the synchronous session starts.**

**Task:** ???

```{r}
# Your code here
```

*Your response here*
